<style type="text/css">
div.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
  text-align: justify;
}
</style>

---
title: "leastcostpath"
author: "Joseph Lewis"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{leastcostpath}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 1. Introducing leastcostpath

The <i>leastcostpath</i> package makes the calculation of conductance surfaces fast and easy. <i>leastcostpath</i> aims to be flexible by multiple time- and energy-based cost functions, as well as allowing for user-supplied functions.

These conductance surfaces can be used to calculate least-cost paths, accumulated cost surfaces, amongst other movement-based outputs.

This documentation introduces the <i>leastcostpath</i> workflow and shows you how to apply it to your data

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 500, scipen = 999)
set.seed(1992)
```

# 2. leastcostpath workflow

```{r libraries, echo = TRUE, message= FALSE, warning= FALSE}
library(sf) # loaded for visualisation purposes
library(terra) # loaded for visualisation purposes
library(leastcostpath)
```

To explore the leastcostpath workflow, we'll use the dataset \code{SICILY_1000m}. This dataset contains an aggregated digital elevation model (1000m) of Sicily. The original dataset (10m resolution) is available here: https://tinitaly.pi.ingv.it/

The digital elevation model (DEM) represents the physical landscape, and forms the foundation for many least-cost path analyses

```{r sicily, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
r <- terra::rast(system.file("extdata/SICILY_1000m.tif", package="leastcostpath"))

r

plot(r)
```

## Creating the ConductanceMatrix

leastcostpath uses a bespoke \code{conductanceMatrix} class. Within this class, information about the supplied digital elevation model (or a spatRaster more generally) is recorded, as well as information specific to the creation of the conductanceMatrix

The implemented cost functions include:
"tobler", "tobler offpath", "davey", 'rees', "irmischer-clarke male", "irmischer-clarke offpath male", "irmischer-clarke female", "irmischer-clarke offpath female", "modified tobler", 'garmy', 'kondo-saino', "wheeled transport", "herzog", "llobera-sluckin", 'naismith', 'minetti', 'campbell', "campbell 2019", "sullivan"

The choosing of which cost function to use is often driven by three considerations: (1) frequently used cost functions, e.g. Tobler's Hiking function (tobler or tobler offpath), (2) whether to use a time-, energy-, or vehicle- based cost function, or (3) cost functions generated from more data and thus deemed to be more representative, e.g. campbell 2019

The implemented cost functions can be divided into:
* Time-based cost functions: "tobler", "tobler offpath", "davey", 'rees', "irmischer-clarke male", "irmischer-clarke offpath male", "irmischer-clarke female", "irmischer-clarke offpath female", "modified tobler", 'garmy', 'kondo-saino', 'campbell', "campbell 2019", "sullivan", 'naismith'
* Energy-based cost functions: "herzog", "llobera-sluckin", 'minetti'
* Vehicle-based cost functions: "wheeled transport"

```{r cm, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
slope_cs <- create_slope_cs(x = r, cost_function = "tobler", neighbours = 4)

slope_cs

plot(slope_cs)

slope_cs_rast <- rasterise(slope_cs)
```

## Understanding the ConductanceMatrix

In the conductanceMatrix, each cell in the original spatRaster is represented by a column. For example, cell 1 is column 1, cell 2 is column 2, etc. 

Each column contains the conductance values for traversing _into_ that cell. For example, column 1, row 2 is accessed via [2,1]. This represents the conductance value from cell 2 into cell 1. If the cells are not traversable (i.e. non-adjacent/not connected), the value in the conductanceMatrix will be 0.

Plotting the slope_cs above, we can see that cell 1 (top-left corner) is NA. As a result, this cell is deemed non-traversable within the conductanceMatrix.

```{r cm1, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
slope_cs$conductanceMatrix[2,1]
head(slope_cs$conductanceMatrix[,1])
quantile(slope_cs$conductanceMatrix[,1])
```

Here, we know that cell 48260 is within the landscape of Sicily. Let's filter to only those that are non-zero (i.e. adjacent/ deemed to be connected).

Given that we used a neighbourhood matrix of 4 when creating the conductanceMatrix, we can see that cell 48260 is connected to 4 different cells.

```{r cm2, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
slope_cs$conductanceMatrix[,48260][slope_cs$conductanceMatrix[,48260] != 0]
```

## Cropping the ConductanceMatrix

If we want to crop the conductanceMatrix to a different extent, we can use the \code{crop_cs} function. For this, we need to supply an sf or terra spatRaster object.

```{r cm3, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
ext <- sf::st_as_sfc(sf::st_bbox(rasterise(slope_cs)))
ext <- sf::st_buffer(ext, dist = -75000)
ext <- sf::st_as_sf(ext)

slope_cs_cropped <- crop_cs(slope_cs, extent = ext)

par(mfrow=c(1,2))

plot(slope_cs, legend = FALSE, main = "before crop")
plot(ext, add = T, border = "red")
plot(slope_cs_cropped, legend = FALSE, main = "after crop")
```

## Modifying values in the ConductanceMatrix

If we want to modify values in the ConductanceMatrix, we can use the \code{update_values} function. 

Here, a function is applied to all conductanceMatrix values that coincide with the supplied sf. Recall that each cell in the spatRaster is represented by each column in the conductanceMatrix. As a result, cells that coincide with the sf object are identified, with these cells as columns used to modify the non-zero values.

Below we modify each conductanceMatrix value that coincides with the sf object by different user-supplied functions

```{r cm4, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
locs4 <- sf::st_sf(geometry = sf::st_sfc(
sf::st_point(c(960745, 4166836)),
crs = terra::crs(r)))

locs4 <- sf::st_buffer(x = locs4, dist = 25000)

slope_cs_mod1 <- slope_cs2 <- update_values(x = slope_cs, sf = locs4, FUN = function(j) { j * 0.6})
slope_cs_mod2 <- slope_cs2 <- update_values(x = slope_cs, sf = locs4, FUN = function(j) { j + 10})
slope_cs_mod3 <- slope_cs2 <- update_values(x = slope_cs, sf = locs4, FUN = function(j) { replace(x = j, values = 0)})

par(mfrow=c(2,2))

plot(slope_cs, legend = FALSE, main = "before mod")
plot(slope_cs_mod1, legend = FALSE, main = "multiply by 0.6")
plot(slope_cs_mod2, legend = FALSE, main = "add 10")
plot(slope_cs_mod3, legend = FALSE, main = "replace with 0")
```

A useful use-case for update_values is the incorporation of estimated speeds attainable when using different modes of transportation.

For example, let us assume that a region of Sicily contains formalised roads. As a result, the estimated speed attainable along these roads is 10km/h (or 2.77778m/s, 10/3.6). We now modify the original \code{slope_cs} to accomodate these roads. Given that update_values allows for all sf objects, linear features can also be incorporated (i.e. not just limited to POLYGON)

To make comparable to other time-based cost functions, we need to account for the resolution of the digital elevation model. We do this by dividing the estimated speed by the maximum resolution (assuming that the spatRaster cells are square)

```{r cm5, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
slope_cs_mod4 <- slope_cs2 <- update_values(x = slope_cs, sf = locs4, FUN = function(j) { replace(x = j, values = (10/3.6)/max(terra::res(r)))})

par(mfrow=c(1,2))

plot(slope_cs, legend = FALSE, main = "before mod")
plot(slope_cs_mod4, legend = FALSE, main = "Incorporating\nestimated speeds")
```

## Replacing values in the ConductanceMatrix using another ConductanceMatrix

Rather than updating values using a function, sometimes we want to replace the conductanceMatrix values using another conductanceMatrix. For this, we can use \code{replace_values}.

A simple use-case should demonstrate:

* We have a time-based (tobler's hiking function) conductanceMatrix
* We have an energy-based (herzog) conductanceMatrix

For whatever reason driven by theory, we want to replace the time-based conductanceMatrix values with energy-based conductanceMatrix values for an area within Sicily. Do note that these two conductanceMatrix have different ranges of value and thus interpretation of conductance values / accumulated cost is less straightforward.

```{r cm6, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
slope_cs2 <- create_slope_cs(x = r, cost_function = "herzog", neighbours = 4)

slope_cs_rep1 <- replace_values(x = slope_cs, y = slope_cs2, sf = locs4)
slope_cs_rep2 <- replace_values(x = slope_cs2, y = slope_cs, sf = locs4)

par(mfrow=c(2,2))

plot(slope_cs, legend = FALSE, main = "time-based")
plot(slope_cs2, legend = FALSE, main = "energy-based")
plot(slope_cs_rep1, legend = FALSE, main = "replaced with energy")
plot(slope_cs_rep2, legend = FALSE, main = "replaced with time")
```

## Forcing conductanceMatrix from anisotropic to isotropic

With some cost functions being anisotropic, i.e. conductance of traversing downhill is different to traversing uphill, the conductance value of moving from cell A to cell B might not be the same as from cell B to cell A. When calculating a least-cost path this therefore means that the least-cost path from an origin to destination might not be same as from destination to origin.

We can force the conductance value to be isotropic, i.e. conductance of traversing downhill is equal to traversing uphill, by taking the mean of the anisotropic conductanceMatrix values from Cell A to Cell B.

```{r cm7, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
slope_cs$conductanceMatrix[48259,48260]
slope_cs$conductanceMatrix[48260,48259]

slope_cs_iso <- force_isotropy(x = slope_cs)

slope_cs_iso$conductanceMatrix[48259,48260]
slope_cs_iso$conductanceMatrix[48260,48259]

mean(c(slope_cs$conductanceMatrix[48259,48260], slope_cs$conductanceMatrix[48260,48259]))

par(mfrow=c(1,2))

plot(slope_cs, legend = FALSE, main = "anisotropic")
plot(slope_cs_iso, legend = FALSE, main = "isotropic")
```

## Creating a user-generated conductanceMatrix

Rather than using a digital elevation model and applying a cost function via \code{create_slope_cs}, we can also create a conductanceMatrix from a user-supplied spatRaster object.

\code{create_cs} assigns the spatRaster value of each cell to each adjacent element in each column within the conductanceMatrix.

```{r cm8, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
locs_8 <- sf::st_sf(geometry = sf::st_sfc(
sf::st_point(c(839769, 4199443)),
crs = terra::crs(r)))

dist_rast <- terra::distance(x = r, y = locs_8[1,])
dist_rast[is.na(r)] <- NA

slope_cs_user <- create_cs(x = dist_rast, neighbours = 4)

par(mfrow=c(1,2))

plot(dist_rast, legend = FALSE, main = "spatRaster")
plot(slope_cs_user, legend = FALSE, main = "conductanceMatrix")

dist_rast[48260]
slope_cs_user$conductanceMatrix[,48260][slope_cs_user$conductanceMatrix[,48260] != 0]
```

## Creating a distance-based conductanceMatrix

Rather than using a digital elevation model and applying a cost function via \code{create_slope_cs}, we can also create a distance-based conductanceMatrix.

The \code{create_distance_cs} uses the supplied spatRaster to retrieve information about the extent, resolution, etc. No spatRaster values are used.

Given a neighbours argument value, the distance conductance from one cell to all other adjacent cells is calculated. 
* If neighbours argument is 4,  the distance conductance value from one cell to all other adjacent cells is equal to the maximum resolution of the supplied spatRaster.
* If neighbours argument is greater than 4, the above values are divided by the distance between adjacent cells to take into account that diagonal distances are further than straight-line distances.

```{r cm9, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
distance_cs_4 <- create_distance_cs(x = r, neighbours = 4)
distance_cs_8 <- create_distance_cs(x = r, neighbours = 8)

par(mfrow=c(1,2))

plot(distance_cs_4, legend = FALSE, main = "4 neighbours")
plot(distance_cs_8, legend = FALSE, main = "8 neighbours")

distance_cs_4$conductanceMatrix[,48260][distance_cs_4$conductanceMatrix[,48260] != 0]
distance_cs_8$conductanceMatrix[,48260][distance_cs_8$conductanceMatrix[,48260] != 0]
```

## Calculating a least-cost path

Using the conductanceMatrix we can calculate a least-cost path, the least costly path from a supplied origin to a supplied destination.

```{r lcp1, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
locs_lcp1 <- sf::st_sf(geometry = sf::st_sfc(
sf::st_point(c(839769, 4199443)),
sf::st_point(c(1038608, 4100024)),
sf::st_point(c(1017819, 4206255)),
sf::st_point(c(1017819, 4206255)),
crs = terra::crs(r)))

lcps <- create_lcp(x = slope_cs, origin = locs_lcp1[1,], destination = locs_lcp1)

plot(r)
plot(locs_lcp1, add = T, pch = 16)
plot(terra::vect(lcps), add = T)
```

## Propagating random error in the DEM to the least-cost path output

It is important to acknowledge that the conductanceMatrix and the least-cost path is derived from a digital elevation model that has random error. Random error here is defined as error that accrues from mistakes during the creation of the digital elevation model.

There are two approaches to deal with random error:
1) We ignore it and hope it doesn't affect our output and our interpretations
2) We incorporate the random error within the modelling process and propagate the error through to the output. That way the output incorporates the error. For more information on this see [Lewis (2021) ](https://doi.org/10.1007/s10816-021-09522-w).

```{r error1, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
locs_lcp2 <- sf::st_sf(geometry = sf::st_sfc(
sf::st_point(c(839769, 4199443)),
sf::st_point(c(1038608, 4100024)),
sf::st_point(c(1017819, 4206255)),
sf::st_point(c(1017819, 4206255)),
crs = terra::crs(r)))

lcps_error <- list()

for(i in 1:10) {

  r_error <- add_dem_error(x = r, rmse = 5, type = "u")
  slope_cs_rand <- create_slope_cs(x = r_error, cost_function = "tobler", neighbours = 4)

  lcps_error[[i]] <- create_lcp(x = slope_cs_rand, origin = locs_lcp2[1,], destination = locs_lcp2)
}

lcps_error <- do.call(rbind, lcps_error)

plot(r)
plot(locs_lcp2, add = T, pch = 16)
plot(terra::vect(lcps_error), add = T, col = "red")
```

## Add global stochasticity in least-cost path output

We can also acknowledge that humans are likely not to have complete knowledge of the topography through which they move. As a result, the least-cost path might not be chosen. We can incorporate this uncertainty by incorporating global stochasticity within the conductanceMatrix. Thus, when we calculate the least-cost path a random component is introduced. See [Pinto and Keitt (2009)](https://doi.org/10.1007/s10980-008-9303-y)  

```{r stoc, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
locs_lcp3 <- sf::st_sf(geometry = sf::st_sfc(
sf::st_point(c(839769, 4199443)),
sf::st_point(c(1038608, 4100024)),
sf::st_point(c(1017819, 4206255)),
sf::st_point(c(1017819, 4206255)),
crs = terra::crs(r)))

lcps_stoch <- list()

for(i in 1:10) {
  
  slope_cs2_stoch <- add_global_stochasticity(slope_cs, percent_quantile = 0.2)

  lcps_stoch[[i]] <- create_lcp(x = slope_cs2_stoch, origin = locs_lcp3[1,], destination = locs_lcp3[2,])
}

lcps_stoch <- do.call(rbind, lcps_stoch)

plot(r)
plot(locs_lcp3[1:2,], add = T, pch = 16)
plot(terra::vect(lcps_stoch), add = T, col = "red")
```

## Calculating a accumulated cost surfaces

we can also calculate the accumulated cost surface from multiple locations. From each supplied location, the accumulated cost to all other cells in the condutanceMatrix is calculated. If more than one location is supplied the calculated accumulated cost surfaces are summarised using the FUN argument. For example, _mean_ will average the values in each pixel of the multiple accumulated cost surfaces

```{r accum, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
locs_accum_cost <- sf::st_sf(geometry = sf::st_sfc(
sf::st_point(c(839769, 4199443)),
sf::st_point(c(1038608, 4100024)),
sf::st_point(c(1017819, 4206255)),
sf::st_point(c(1017819, 4206255)),
crs = terra::crs(r)))

accum_cost <- create_accum_cost(x = slope_cs, origins = locs_accum_cost, FUN = mean, rescale = FALSE)

plot(accum_cost)
plot(locs_accum_cost, add = T, pch = 16)
```

## Calculating a cost corridor

If we want to understand movement between two locations, we can calculate a cost corridor. This calculates the accumulated cost from the origin and the accumulated cost from the destination. These two accumulated cost surfaces are averaged resulting in a final surface that represents the average cost when moving from origin to destination and destination to origin. From here, we can filter to those costs that are below a subjectively defined threshold to identify a 'cost corridor'

```{r cc, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
locs_cc <- sf::st_sf(geometry = sf::st_sfc(
sf::st_point(c(839769, 4199443)),
sf::st_point(c(1038608, 4100024)),
crs = terra::crs(r)))

cost_corridor <- create_cost_corridor(x = slope_cs, origin = locs_cc[1,], destination = locs_cc[2,], rescale = TRUE)

par(mfrow=c(1,2))

plot(cost_corridor)
plot(locs_cc, add = T, pch = 16)

plot(cost_corridor < 0.01, main = "Lowest 1% Cost")
plot(locs_cc, add = T, pch = 16)
```

## Calculate From-Everywhere-to-Everywhere Least-cost paths

In many cases we don't know the origins and destinations to connect. We can overcome this by calculating least-cost paths between multiple origins and destinations. See [White and Barber (2012)](https://doi.org/10.1016/j.jas.2012.04.017)

```{r fete, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
locs_fete <- sf::st_sf(geometry = sf::st_sfc(
sf::st_point(c(839769, 4199443)),
sf::st_point(c(1038608, 4100024)),
sf::st_point(c(907695, 4145478)),
sf::st_point(c(1044758, 4215739)),
crs = terra::crs(r)))

lcps_fete <- create_FETE_lcps(x = slope_cs, locations = locs_fete)

plot(r)
plot(locs_fete, add = T, pch = 16)
plot(lcps_fete, add = T, col = "red")
```

## Creating a least-cost path density surface

We can identify pixels that have higher/fewer numbers of least-cost paths by creating the density of lines


```{r dens, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
lcps_dens <- create_lcp_density(r, lcps_fete)

plot(lcps_dens)
plot(locs_fete, add = T, pch = 16)
```

## Validate a least-cost path using the buffer method

After calculating a least-cost path we might want to validate it against a route that we're trying to model. One approach is using the buffer method as proposed by [Goodchild and Hunter (1997)](https://doi.org/10.1080/136588197242419). Here the percentage of the least-cost path within a supplied buffer from the comparison route is returned

```{r val1, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
locs_val_buffer <- sf::st_sf(geometry = sf::st_sfc(
sf::st_point(c(839769, 4199443)),
sf::st_point(c(1038608, 4100024)),
crs = terra::crs(r)))

lcp_val_buffer1 <- create_lcp(x = slope_cs, origin = locs_val_buffer[1,], destination = locs_val_buffer[2,])
lcp_val_buffer2 <- create_lcp(x = slope_cs, origin = locs_val_buffer[2,], destination = locs_val_buffer[1,])

plot(r)
plot(locs_val_buffer[1:2,], add = T, pch = 16)
plot(terra::vect(lcp_val_buffer1), add = T, col = "red")
plot(terra::vect(lcp_val_buffer2), add = T, col = "blue")

buffer_validation(lcp = lcp_val_buffer1, comparison = lcp_val_buffer2, dist = c(1000, 3000, 5000, 10000))
```

## Validate a least-cost path using the path deviation index method

An alternative validation method is the Path Deviation Index (PDI). Here the spatial separation between a pair of paths is calculated. This method aims to overcome the shortcomings of measuring the percentage of coverage of a least cost path from a comparison path

The pdi index is defined as the area between paths divided by the Euclidean distance of the shortest path between the origin and destination of the paths. The index can be interpreted as the average distance between the paths.

```{r val2, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
locs_val_pdi <- sf::st_sf(geometry = sf::st_sfc(
sf::st_point(c(839769, 4199443)),
sf::st_point(c(1038608, 4100024)),
crs = terra::crs(r)))

lcp_val_pdi1 <- create_lcp(x = slope_cs, origin = locs_val_pdi[1,], destination = locs_val_pdi[2,])
lcp_val_pdi2 <- create_lcp(x = slope_cs, origin = locs_val_pdi[2,], destination = locs_val_pdi[1,])

lcp_diff <- PDI_validation(lcp = lcp_val_pdi1, comparison = lcp_val_pdi2)

lcp_diff

plot(r)
plot(lcp_diff, add = T, col = "black")
plot(locs_val_pdi[1:2,], add = T, pch = 16)
plot(terra::vect(lcp_val_pdi1), add = T, col = "red")
plot(terra::vect(lcp_val_pdi2), add = T, col = "blue")
```
