<style type="text/css">
div.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
  text-align: justify;
}
</style>

---
title: "leastcostpath-1"
author: "Joseph Lewis"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: false
vignette: >
  %\VignetteIndexEntry{leastcostpath-1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 1. Introducing leastcostpath

This vignette describes <b>leastcostpath</b>, a package written for use in the R environment (R Core Team, 2018) and built on classes and functions provided in the R package gdistance (Van Etten, 2017).

<b>leastcostpath</b> provides the functionality to calculate Least Cost Paths (LCPs) using numerous time- and energy-based cost functions that approximate the difficulty of moving across a landscape. Additional cost surfaces can be incorporated into the analysis via create_barrier_cs() or create_feature_cs().

<b>leastcostpath</b> provides the functionality to calculate Stochastic Least Cost Paths (Pinto and Keitt, 2009), and Probabilistic Least Cost Paths (Lewis, 2020). 

<b>leastcostpath</b> provides the functionality to calculate movement potential within a landscape through the implementation of From-Everywhere-to-Everywhere (White and Barber, 2012), Cumulative Cost Paths (Verhagen 2013), and Least Cost Path calculation within specified distance bands (Llobera, 2015).

<b>leastcostpath</b> provides the functionality to validate the accuracy of the computed Least Cost Path relative to another path via validate_lcp() (Goodchild and Hunter, 1997) and PDI_validation() (Jan et al. 1999).

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 500)
```

# 2. using leastcostpath

## 2.1 Setup

```{r libraries, echo = TRUE, message= FALSE, warning= FALSE}
library(rgdal)
library(rgeos)
library(sp)
library(raster)
library(Matrix)
library(gdistance)
library(leastcostpath)
```

```{r raster, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
dem <- raster::raster(system.file('external/maungawhau.grd', package = 'gdistance'))

raster::crs(dem) <- sp::CRS("+init=epsg:27200")

dem_extent <- as(raster::extent(dem), 'SpatialPolygons')

raster::crs(dem_extent) <- raster::crs(dem)

raster::plot(dem)
raster::plot(dem_extent, add = T, border = "red")
```

## 2.2 Creating Slope-based Cost Surfaces

```{r cs, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
# cost functions currently implemented within leastcostpath

cfs <- c("tobler", "tobler offpath", "irmischer-clarke male", 
         "irmischer-clarke offpath male", "irmischer-clarke female", 
         "irmischer-clarke offpath female","modified tobler", 
         "wheeled transport", "herzog", "llobera-sluckin", "campbell 2019")

# neighbours can be 4, 8, 16, 32, or 48. 4 used for illustration purposes, but greater number of neighbours = cost surface and LCP approximates reality better.

neigh <- 4

slope_cs <- leastcostpath::create_slope_cs(dem = dem, cost_function = "tobler", neighbours = neigh)

plot(raster(slope_cs), col = grey.colors(100))
```

## 2.3 Creating and Incorporating Barriers within Cost Surfaces

```{r barrier, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
# create barrier of altitude values above 180

altitude <- dem > 180
# convert value 0 to NA - this ensures that the areas below 180 are not viewed as barriers. That is, if the raster value is NA then create_barrier_cs() will assume that the cell is background and will therefore assign the background argument value 
altitude[altitude == 0] <- NA

# the values NOT NA will be assigned the field argument value. If 0 (default) then movement within the area will be completely prohibited
altitude_cs <- leastcostpath::create_barrier_cs(raster = altitude, barrier = altitude, neighbours = neigh, field = 0, background = 1)

# multiplying the two cost surfaces ensures that barriers continue to completely inhibit movement (i.e. slope_cs values * 0 = 0)
slope_altitude_cs <- slope_cs * altitude_cs

plot(raster(slope_altitude_cs), col = grey.colors(100))
```

## 2.3 Calculating Least Cost Paths

```{r lcp = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
A <- sp::SpatialPoints(cbind(2667930, 6479466))
B <- sp::SpatialPoints(cbind(2667668, 6478818))

# if cost function is anisotropic (e.g. Tobler's Hiking Function) then LCP from A-to-B and B-to-A may be different. To create LCP in both directions set the directional argument to FALSE (default)
lcp <- leastcostpath::create_lcp(cost_surface = slope_altitude_cs, origin = A, destination = B, directional = FALSE)

plot(raster(slope_altitude_cs), col = grey.colors(100))
plot(A, add = T, col = "black")
plot(B, add = T, col = "black")
# LCP from A-to-B
plot(lcp[1,], add = T, col = "red")
# LCP from B-to-A
plot(lcp[2,], add = T, col = "blue")
```

## 2.4 Validating Least Cost Paths

```{r validation = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
lcp_A <- sp::SpatialPoints(cbind(2667930, 6479466))
lcp_B <- sp::SpatialPoints(cbind(2667592, 6478854))

lcp <- leastcostpath::create_lcp(cost_surface = slope_altitude_cs, origin = lcp_A, destination = lcp_B, directional = TRUE)

comparison_A <- sp::SpatialPoints(cbind(2667930, 6479466))
comparison_B <- sp::SpatialPoints(cbind(2667592, 6478854))

comparison <- leastcostpath::create_lcp(cost_surface = slope_cs, origin = comparison_A, destination = comparison_B, directional = TRUE)

validate_lcp(lcp = lcp, comparison = comparison, buffers = c(1, 10, 100))
PDI <- PDI_validation(lcp = lcp, comparison = comparison)
data.frame(PDI@data)

plot(raster(slope_altitude_cs), col = grey.colors(100))
plot(lcp_A, add = T, col = "black")
plot(lcp_B, add = T, col = "black")

plot(lcp, add = T, col = "red")
plot(comparison, add = T, col = "blue")
plot(PDI, add = T, col = rgb(red = 0, green = 1, blue = 0, alpha = 0.4))
```

## 2.5 Calculating Cost Corridors

```{r cc, TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
# calculates accumulated cost surface from A and from B and B to A. These two cost surfaces are averaged. If rescale argument TRUE (not default) then final accumulated cost surface rescaled to 1. 

cc <- leastcostpath::create_cost_corridor(cost_surface = slope_altitude_cs, origin = A, destination = B, rescale = TRUE)

plot(cc, col = heat.colors(100))
plot(A, add = T, col = "white", pch = 16)
plot(B, add = T, col = "white", pch = 16)
```

## 2.6 Calculating From-Everywhere-To-Everywhere Least Cost Paths

```{r fete, TRUE,  fig.height = 6, fig.width = 6, message=FALSE, warning=FALSE}
set.seed(1992)
random_locs <- sp::spsample(x = dem_extent, n = 5, type = 'regular')

# if parallel is TRUE then need to specify number of cores via ncores argument 
fete_lcp <- leastcostpath::create_FETE_lcps(cost_surface = slope_altitude_cs, locations = random_locs, cost_distance = FALSE, parallel = FALSE)

plot(raster(slope_altitude_cs), col = grey.colors(100))
plot(fete_lcp, add = T, col = "red")
plot(random_locs, add = T, col = "black", pch = 16)
```

## 2.7 Least Cost Path Density

```{r lcp_d, echo = TRUE,  fig.height = 6, fig.width = 6, warning = FALSE}
# rasterises the LCPs and calculates the number of times an LCP crosses a raster cell. If rescale argument TRUE (not default) then LCP density rescaled to 1. If rasterize_as_points argument TRUE (default) then LCP vertices points are rasterized. This is quicker than rasterizing lines (rasterize_as_lines argument FALSE) but will contain 'gaps' where no LCP vertex point is present 
fete_lcp_density <- leastcostpath::create_lcp_density(lcps = fete_lcp, raster = dem, rescale = TRUE, rasterize_as_points = TRUE)

fete_lcp_density[fete_lcp_density == 0] <- NA

plot(fete_lcp_density, col = heat.colors(100))
```

# References

Goodchild, M. F., & Hunter, G. J. (1997). A simple positional accuracy measure for linear features. International Journal of Geographical Information Science, 11(3), 299–306. https://doi.org/10.1080/136588197242419

Jan, O., Horowitz, A., & Peng, Z. (1999). Using GPS Data to Understand Variations in Path Choice.

Llobera, M. (2015). Working the digital: some thoughts from landscape archaeology. Material evidence: learning from archaeological practice. Routledge, Abingdon, 173–188.

Pinto, N., & Keitt, T. H. (2009). Beyond the least-cost path: evaluating corridor redundancy using a graph-theoretic approach. Landscape Ecology, 24(2), 253–266. https://doi.org/10.1007/s10980-008-9303-y

R Core Team. (2018). R: A language and environment for statistical computing. Viena, Austria: R Foundation for Statistical Computing. https://www.R-project.org/

van Etten, J. (2017). R Package gdistance: Distances and Routes on Geographical Grids. Journal of Statistical Software, 76(13). https://doi.org/10.18637/jss.v076.i13

Verhagen, P. (2013). On the Road to Nowhere?: Least Cost Paths, Accessibility and the Predictive Modelling Perspective (pp. 383–389). Presented at the CAA2010: fusion of cultures: proceedings of the 38th Annual Conference on Computer Applications and Quantitative Methods in Archaeology, Granada, Spain, April 2010, Archaeopress.

White, D., & Barber, S. L. (2012). Geospatial modeling of pedestrian transportation networks: a case study from precolumbian Oaxaca, Mexico. Journal of Archaeological Science, 39(8), 2684–2696. https://doi.org/10.1016/j.jas.2012.04.017
